---
title: "R Notebook"
output: html_notebook
---

```{r requirements, include=F}
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(scales)
```

Data were extracted from the [OpenFDA adverse events](https://open.fda.gov/drug/event/) using `fetch_2015.R`. This extracts a subset of the data which was recieved in 2015.

```{r read_data}
reports <- read_csv('data/fda/raw/2015_reports.csv', col_types = cols(
  reportid = col_character(),
  primarysource = readr::col_factor(levels = c("Physician", "Pharmacist","Other health professional", "Lawyer", "Consumer or non-health professional", "Unknown")),
  patientage = col_integer(),
  patientageunit = readr::col_factor(levels = c("Decade", "Year", "Month", "Week", "Day", "Hour")),
  patientsex = readr::col_factor(c("Unknown", "Male", "Female")),
  country = readr::col_factor(levels = NULL),
  transmission_date = col_date("%Y-%m-%d"),
  receipt_date = col_date("%Y-%m-%d")
)
) %>% mutate(
  country = fct_infreq(country)
)
  
drugs   <- read_csv('data/fda/raw/2015_drugs.csv', col_types = cols(
  reportid = col_character(),
  Drug = col_character(),
  Indication = col_character()
))
reacts  <- read_csv('data/fda/raw/2015_reacts.csv', col_types = cols(
  reportid = col_character(),
  React = col_character()
))
```

There are `r nrow(reports)` in the dataset, which represents `r reacts %>% select(React) %>% unique %>% nrow()` types of adverse event, from `r drugs %>% select(Drug) %>% unique %>% nrow()` drugs.


```{r reports_by_gender}
reports %>% count(patientsex) %>% arrange(-n) %>% ggplot(aes(x = patientsex, y = n)) +
  geom_point() + labs(y = "# of reports", x = "Patient Sex") 
```

Almost twice as many adverse event reports are concerning women.

## Differences by country

There's a big bias in the data toward the US, with almost 2/3 of the events coming from there, so any comparator should be normalized by the number of reports per country.

```{r reports_per_country}
reports_per_country <- reports %>% count(country) %>% arrange(-n)
reports_per_country %>% head(10)
```


```{r primary_source}
reports %>% ggplot(aes(x = primarysource)) + geom_bar() + labs(x = 'Reporter of AE', y = 'count') + scale_y_continuous(labels = comma) + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0)) +
   scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 15))
```


Most events are reported by the consumer themselves, followed by physians and then other healthcare professionsals.  Could be interesting to look at this over time and see how that changed (if at all).  Very few lawyers are involved in the process it seems. 

```{r primary_source_by_country, fig.width=8}
# do this long-hand as there doesn't seem to be a nice ggplot option to scale the output weel
country_subset <- reports %>%
  mutate(country = fct_lump(country, n = 11)) %>%
  filter(!is.na(country)) # this only represents 168 events

events_per_country <- country_subset %>% group_by(country) %>% count() %>% rename(total_per_country = n)

country_subset %>%
  group_by(country, primarysource) %>%
  count() %>%
  left_join(events_per_country, by = 'country') %>%
  mutate(n = n/total_per_country) %>%
  ggplot(aes(x = primarysource, y = n*100)) + geom_bar(stat= 'identity') + labs(x = 'Reporter of AE', y = '% from source') + scale_y_continuous(labels = comma) + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0)) +
   scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 20)) + facet_wrap(~country)
```

Unsuprisingly, the US shows the same shape as the overall picture.  The already very rare laywers don't seem to feature at all in reports outside the US. Canada has a similar picture to the US, though more balanced between Health professionals and consumer reports.  Brazil is skewed the other way, with most reports coming from non-health professionals.  Japan and Germany show a relatively strong lean toward physicians. Generally the rest of the world seems to have a lean toward events being reported from healthcare professionals -- is this a reflection of how things are reported, or are healthcare professional reports more likely to be cascaded to the FDA?

## Adverse Events distribution

Top 10 adverse events per report

```{r top10_adverse_events}
reaction_counts <- reacts %>% unique() %>% count(React) %>% arrange(-n)
reaction_counts %>% head(10)
```



```{r reactions_per_report}
reports %>% left_join(reacts %>% unique %>% count(reportid)) %>% ggplot(aes(x = n, colour = patientsex)) + geom_freqpoly(bins = 50) + labs(x = "Number of events per report", title = "Distribution of number of adverse events on one report")
```


## drugs distribution

Top 10 drugs per report, treating multiple occurences on one report as 1 appearance

```{r top10_drugs_events}
drugs %>% select(reportid, Drug) %>% unique() %>% count(Drug) %>% arrange(-n) %>% head(10)
```

```{r drug_per_report}
reports %>% left_join(drugs %>% select(reportid,Drug) %>% unique %>% count(reportid)) %>% ggplot(aes(x = n, colour = patientsex)) + geom_freqpoly(bins = 50) + labs(x = "Number of drugs per report", title = "Distribution of number of drugs on one report")
```

The vast bulk of reports have one or two drugs involved.

There's no obvious relationship between number of reported reactions and the number of drugs taken

```{r reacts_and_drugs}
reports %>%
  left_join(reacts %>% unique %>% count(reportid) %>% rename(nreacts = n)) %>%
  left_join(drugs %>% select(reportid,Drug) %>% unique %>% count(reportid) %>% rename(ndrugs = n)) -> annotated_reports

ggplot(annotated_reports, aes(x = ndrugs, y = nreacts)) + geom_jitter(alpha = 0.1)
```


## By Age

Distribution of reports by patient age

```{r by_age}
reports %>% filter(patientageunit == 'Year', !is.na(patientage)) %>% count(patientage) %>% ggplot(aes(x = patientage, y = n)) + geom_point()
```





Can we see any difference in the types of reports by country?

```{r prepare_reports}
common_reactions <- reaction_counts %>%
  filter(n >= nrow(reports)*0.001) %>% # reactions that occur in at least 0.1% of reports
  semi_join(reacts, ., by = 'React') # select only those reactions

reports_with_reactions <- reports %>%
  inner_join(common_reactions %>% unique(), by = 'reportid')

reports_per_country_subset <- reports_with_reactions %>% count(country) %>% rename(n_per_country = n)

# construct the report signature for each country
report_signature <- reports_with_reactions %>%
  select(reportid, country, React) %>%
  left_join(reports_per_country_subset, by = 'country') %>%
  filter(n_per_country >= 500) %>% # pick only countries with at least 500 reports
  mutate(value = 1) %>%
  spread(React, value, fill = 0) %>%
  select(-reportid, -n_per_country) %>%
  group_by(country) %>%
  summarise_all(sum) %>%
  left_join(reports_per_country_subset, by = 'country') %>%
  mutate_if(is.double, funs(. / n_per_country)) %>%
  select(-n_per_country) 
```

```{r clustering}

report_matrix <- report_signature %>% select(-country) %>% as.matrix() 
rownames(report_matrix) <- report_signature %>% select(country) %>% unlist()

d <- dist(report_matrix, d = 'can')

hc <- hclust(d, method ='ward.D')
plot(hc)
nGroup <- 5
groups <- cutree(hc, k=nGroup) # cut tree into 5 clusters
rect.hclust(hc, k=nGroup, border="red")
```

```{r cluster_mds}
md <- cmdscale(d) %>%
  as.data.frame() %>%
  rownames_to_column(var = 'country') %>%
  mutate(group = groups)

ggplot(md, aes(x = V1, y = V2, colour = as.factor(group))) + geom_point() + geom_text(aes(label = country), hjust = 0.1, vjust = 0.1) + scale_color_brewer(palette = 'Dark2') + theme_bw() + labs(x = '', y = '', colour = 'Cluster', title = "MDS plot of the countries clustered on, coloured by group")

```

```{r clustering2}

report_matrix <- report_signature %>% select(-country) %>% as.matrix() 
rownames(report_matrix) <- report_signature %>% select(country) %>% unlist()

d <- dist(report_matrix, d = 'can')

hc <- hclust(d, method ='ward.D')
plot(hc)
nGroup <- 10
groups <- cutree(hc, k=nGroup) # cut tree into 5 clusters
rect.hclust(hc, k=nGroup, border="red")
```
